GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations

### 2022-12-19 writeUps/paper.tex

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{amsmath,amssymb}
\usepackage{bm}
\usepackage{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{fullpage}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{numdef}

@@

@@

Because principal strata membership is generally unobserved, principal effects are not always identified, and when they are, estimation can still be difficult.
Typically, principal effects are identified and estimated based on strong structural assumptions \citep[e.g.][]{air} and/or parametric modeling \citep[e.g.][]{imbens1997bayesian}.
Parametric models for principal effects are often hard to specify and hard or impossible to test.
Even when they are well-specified they can yield estimates that are severely biased and misleading \citep{griffin2008application,feller2016principal}.
Another route to identification and estimation of principal effects leads through auxiliary data, including baseline covariates or secondary outcomes \citep{mattei2013exploiting}.
Of particular importance is the ``principal score,'' \citep{jo,ding2017principal,feller2017principal}, or the probability that a subject belongs to a particular principal stratum, as a function of pre-treatment covariates.
Under a strong, untestable assumption called ``principal ignorability,'' which is akin to the no-confounding assumption underlying observational studies, principal scores can be transformed into weights, which in turn form an unbiased estimate of principal effects.

@@

The following section gives a formal introduction to principal stratification and briefly describes two common estimation techniques, parametric mixture modeling and principal score weighting.
Next, section \ref{sec:geepers} develops GEEPERs estimation, and gives conditions for its consistency.
Section \ref{sec:simulation} describes a simulation study examining the operating characteristics of GEEPERs estimates and comparing them to estimates from parametric mixture modeling and principal score weighting.
Under ideal circumstances, estimates from parametric mixture models and weighting estimators are somewhat more efficient than GEEPERs, but when assumptions are violated GEEPERs easily out-performs the other two methods.
Section \ref{sec:fh2t} illustrates GEEPERs in analyzing data from a recent large-scale field trial of educational software, and Section \ref{sec:discussion} concludes.

@@

Let $S_i\in\{0,1\}$ be a measurement on subject $i$ taken subsequent to treatment assignment, so that $S$ may be affected by $Z$.
Then $S$ itself has potential values $\sti$ and $\sci$, which $i$ would exibit if $Z_i=1$ or $Z_i=0$, respectively.
Assume ``strong monotonicity'' \citep[c.f.][]{dingLu} (also referred to as ``one-way noncompliance'' \citealt{wing2017can}):
\begin{ass}[Strong Monotonicity]\label{ass:sm}
\begin{equation*}%\label{eq:strongMonotonicity}
  \sci=0\text{ for all }i
\end{equation*}
\end{ass}
Like potential outcomes $\yti$ and $\yci$, $\sci$ and $\sti$ are defined for all subjects, though $\sci$ is only observed when $Z_i=0$ and $\sti$ is only observed when $Z_i=1$. Under strong monotonicity, since $S$ is binary, every subject belongs to one of two ``principal strata,'' $\sti=1$ or $\sti=0$.
The average treatment effect in each principal stratum, $\eff1=\EE[\tau|\st=1]$ or $\eff1=\EE[\tau|\st=0]$, is called a principal effect. %
The challenge in estimating principal effects is due to the fact that strata membership $\sti$ is only observed when $Z_i=1$, in which case $\sti=S_i$, the observed value of $S$.

@@

\subsection{Principal Scores}
The principal score \citep[e.g.][]{jo2009use} is defined as
\begin{equation}\label{eq:pscore}
  \ppi=Pr(\sti=1|\bxsi)
\end{equation}
where $\bxsi\subseteq\bm{x}_i$, the set of covariates used to model $S$.
Note that under randomization, $Pr(\sti=1|\bxsi,Z_i)=Pr(\sti=1|\bxsi)$, so a model for principal scores can be estimated using data from the treatment group and extrapolated to the control group.
Assume a model for principal scores:
\begin{equation}\label{eq:prinScoreMod}
  \ppip{\bm{\alpha}}=f(\bxsi;\bm{\alpha})
\end{equation}
with parameter vector $\bm{\alpha}$. An analyst may estimate $\bm{\alpha}$ by fitting model \eqref{eq:prinScoreMod} using observed $\bm{S}$ and $\bxs$ for subjects with $Z=1$, and then compute estimated principal scores $\ppip{\bm{\hat{\alpha}}}$ for subjects with $Z_i=1$.

@@

Alternatively, an analyst may assume ``principal ignorability'' \citep{jo2009use,dingLu}:
\begin{ass}[Principal Ignorability]\label{ass:PI}
\begin{equation}\label{eq:pi}
  Y_{C}\independent \st |\bxs
\end{equation}
\end{ass}
or a somewhat weaker version, $\EE[Y_C|\st,\bxs]=\EE[Y_C|\bxs]$ \citep{feller2017principal}.
That is, the principal stratum is unrelated to control potential outcomes conditional on covariates.
Principal ignorability \eqref{eq:pi} is reminiscent of ignorability assumptions typical in observational studies \citep[e.g.][]{rosenbaum2010design}.
% Both require a rich set of covariates and background knowledge of the determinants of $S$; 
In particular, an unobserved covariate $u$ that is correlated with both $\st$ and $Y_C$ would invalidate \eqref{eq:pi}.

@@

Principal score weighting does not require any distributional assumptions and (after estimating principal scores) very easy to implement.
On the other hand, the principal ignorability assumption is strong and restrictive.

\citet{feller2017principal} recommends the case-resampling bootstrap to estimate the sampling variance of $\hat{\muc0}_w$, $\hat{\muc1}_w$, and their associated principal effect estimators.

@@

We conducted a simulation study to compare the performance of GEEPERs compared to a Bayesian mixture model and to a principal score weighting estimator, in both favorable and unfavorable circumstances.
The study was designed to answer three sets of overarching questions.
First, how does the performance of the GEEPERs estimator vary under different conditions, including sample size, the extent to which $\bx$ predicts $\st$, the principal effects themselves vary, and when important interactions were omitted from the outcome model.
Second, how does GEEPERs compare with mixture modeling? Is it competitive in circumstances favorable to both techniques? Does it avoid the pitfalls of mixture modeling when the parametric assumptions of the normal mixture model fail?
Third, are there ways or circumstances when the principal score weighting estimator outperforms GEEPERs even though principal ignorability does not hold?

@@

In each run of the simulation, we simulated three independent standard normal covariates, $x_k$ $k=1,\dots,3$. However, only the first two covariates were ``observed,'' i.e. were included in the analysis model.
Given the covariates, the (true) principal scores were defined as
\begin{equation*}
  \ppi=logit^{-1}\left[\alpha(x_{1i}-x_{2i}+x_{3i})\right]
\end{equation*}
where $\alpha$ is a manipulated factor.
When $\alpha$ was higher, $\st$ was more easily predicted by covariates; when $\alpha=0$, Assumption \ref{ass:vps}, was violated, i.e. $Var(\pp)=0$.
Then principal stratum $\sti\in \{0,1\}$ was simulated as $S_i\sim Bern(\ppi)$

Finally, potential and observed outcomes were generated as in \eqref{eq:interaction}, with $\bxy=[\bm{x_1},\bm{x_2},\bm{x_3}]$,
and with $\EE[\epsilon_i]=0$ and $Var(\epsilon_i)=1/2$.
Specifically,
\begin{equation}\label{eq:y-sim}
Y_i=\beta_1\sti+\beta_3Z_i\sti+(\gamma_1+\gamma_2\sti+\gamma_3Z_i)(x_{1i}+x_{2i})+\frac{1}{\sqrt{6}}x_{3i}+\epsilon_i%\\
\end{equation}
%   \begin{split}
%   \yci&=\beta_1\sti+(\gamma_1+\gamma_2\sti)(x_{1i}+x_{2i})+\frac{1}{\sqrt{6}}x_{3i}+\epsilon_i\\
%   \yti&=0.3\sti+(\gamma_1+\gamma_2\sti+\gamma_3Z_i)(x_{1i}+x_{2i})+\frac{1}{\sqrt{6}}x_{3i}+\epsilon_i\\
%   Y_i&=Z_i\yti+(1-Z_i)\yci
%   \end{split}
% \end{equation}
with $\beta_3=0.3-\beta_1$. Hence, $\beta_0=\beta_2=\gamma_4=0$, when $\gamma_3=0$ (i.e. no interaction between $Z$ and covariates), there is no average treatment effect in the $\st=0$ stratum, the coefficient for the unobserved covariate $x_3$ is constant across treatment groups and principal strata, and we do not consider three-way interactions between $x_1$ or $x_2$, $\st$, and $Z$.

@@

\textbf{\emph{Sample Size per condition $\bm{n}$:}} The guarantees of M-estimation are all asymptotic, so we examined the behavior of our M-estimator at a range of sample sizes to determine its finite-sample properties.

@@

Following \citet[][p. 373]{carroll2006measurement}, we %separate the estimating equations $\Psi$ into $\phi$ and $\psi$, estimating equations for models \eqref{
can decompose the matrices into diagonal elements 
\begin{equation*}
    \begin{split}
        A_{1,1}&=\sum_i \partial \Omega_i/\partial \bm{\alpha}|_{\bm{\alpha}=\bm{\hat{\alpha}}}\\
        A_{2,2}&=\sum_i\partial\Psi_i/\partial \bm{\beta}|_{\bm{\beta}=\bm{\hat{\beta}}}\\
        B_{1,1}&=\sum_i\Omega_i(\bm{\hat\alpha})\Omega_i(\bm{\hat\alpha})'\\
        B_{2,2}&=\sum_i \Psi_i(\bm{\hat\alpha},\bm{\hat\beta})\Psi_i(\bm{\hat\alpha},\bm{\hat\beta})'
    \end{split}
\end{equation*}
 that pertain to the parameter sets $\bm{\alpha}$ and $\bm{\beta}$ and the estimating equations for models \eqref{eq:psMod} and  \eqref{eq:regression}, respectively, and 
 \begin{equation*}
     \begin{split}
         A_{21}&=\sum_i\partial\Psi_i/\partial \bm{\alpha}|_{\bm{\alpha}=\bm{\hat{\alpha}}}\\
         B_{12}=B_{21}'&=\sum_i \Omega_i(\bm{\hat{\alpha}})\Psi_i(\bm{\hat{\alpha}},\bm{\hat{\beta}})'
     \end{split}
 \end{equation*}
 which capture to the dependence of model \eqref{eq:regression} on the parameters $\bm{\alpha}$ from \eqref{eq:psMod} and the covariance between the estimating equations of the two models. 
The sub-matrix $A_{12}=\sum_i \partial \Omega_i/\partial \bm{\beta}=0$, since \eqref{eq:psMod} does not depend on $\bm{\beta}$.

@@

>>>>

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{amsmath,amssymb}
\usepackage{bm}
\usepackage{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{fullpage}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{numdef}
\usepackage{multibib}

@@

\newcites{supp}{Supplementary References}

@@

Because principal strata membership is generally unobserved, principal effects are not always identified, and when they are, estimation can still be difficult.
Typically, principal effects are identified and estimated based on strong structural assumptions \citep[e.g.][]{air} and/or parametric modeling \citep[e.g.][]{imbens1997bayesian}.
Parametric models for principal effects are often hard to specify and hard or impossible to test.
Even when they are well-specified they can yield estimates that are severely biased and misleading \citep{griffin2008application,feller2016principal}.
Another route to identification and estimation of principal effects leads through auxiliary data, including baseline covariates or secondary outcomes \citep{mattei2013exploiting}.
Of particular importance is the ``principal score,'' \citep{jo,dingLu,feller2017principal}, or the probability that a subject belongs to a particular principal stratum, as a function of pre-treatment covariates.
Under a strong, untestable assumption called ``principal ignorability,''  principal scores can be transformed into weights to compute form an unbiased estimate of principal effects.

@@

The following section gives a formal introduction to principal stratification and briefly describes two common estimation techniques, parametric mixture modeling and principal score weighting.
Next, Section \ref{sec:geepers} develops GEEPERs estimation, and gives conditions for its consistency.
Section \ref{sec:simulation} describes a simulation study examining the operating characteristics of GEEPERs estimates and comparing them to estimates from parametric mixture modeling and principal score weighting.
Under ideal circumstances, estimates from parametric mixture models and weighting estimators are somewhat more efficient than GEEPERs, but when assumptions are violated GEEPERs easily out-performs the other two methods.
Section \ref{sec:fh2t} illustrates GEEPERs in analyzing data from a recent large-scale field trial of educational software, and Section \ref{sec:discussion} concludes.

@@

Let $S_i\in\{0,1\}$ be a measurement on subject $i$ taken subsequent to treatment assignment, so that $S$ may be affected by $Z$.
Then $S$ itself has potential values $\sti$ and $\sci$, which $i$ would exibit if $Z_i=1$ or $Z_i=0$, respectively.
Assume ``strong monotonicity'' \citep[c.f.][]{dingLu} (also referred to as ``one-way noncompliance''):%\citealt{wing2017can}):
\begin{ass}[Strong Monotonicity]\label{ass:sm}
\begin{equation*}%\label{eq:strongMonotonicity}
  \sci=0\text{ for all }i
\end{equation*}
\end{ass}
Like potential outcomes $\yti$ and $\yci$, $\sci$ and $\sti$ are defined for all subjects, though $\sci$ is only observed when $Z_i=0$ and $\sti$ is only observed when $Z_i=1$. Under strong monotonicity, since $S$ is binary, every subject belongs to one of two ``principal strata,'' $\sti=1$ or $\sti=0$.
The average treatment effect in each principal stratum, $\eff1=\EE[\tau|\st=1]$ or $\eff1=\EE[\tau|\st=0]$, is called a principal effect. %
The challenge in estimating principal effects is due to the fact that strata membership $\sti$ is only observed when $Z_i=1$, in which case $\sti=S_i$, the observed value of $S$.

@@

\subsection{Principal Scores}
The principal score \citep[e.g.][]{jo} is defined as
\begin{equation}\label{eq:pscore}
  \ppi=Pr(\sti=1|\bxsi)
\end{equation}
where $\bxsi\subseteq\bm{x}_i$, the set of covariates used to model $S$.
Note that under randomization, $Pr(\sti=1|\bxsi,Z_i)=Pr(\sti=1|\bxsi)$, so a model for principal scores can be estimated using data from the treatment group and extrapolated to the control group.
Assume a model for principal scores:
\begin{equation}\label{eq:prinScoreMod}
  \ppip{\bm{\alpha}}=f(\bxsi;\bm{\alpha})
\end{equation}
with parameter vector $\bm{\alpha}$. An analyst may estimate $\bm{\alpha}$ by fitting model \eqref{eq:prinScoreMod} using observed $\bm{S}$ and $\bxs$ for subjects with $Z=1$, and then compute estimated principal scores $\ppip{\bm{\hat{\alpha}}}$ for subjects with $Z_i=1$.

@@

Alternatively, an analyst may assume ``principal ignorability'' \citep{jo,dingLu}:
\begin{ass}[Principal Ignorability]\label{ass:PI}
\begin{equation}\label{eq:pi}
  Y_{C}\independent \st |\bxs
\end{equation}
\end{ass}
or a somewhat weaker version, $\EE[Y_C|\st,\bxs]=\EE[Y_C|\bxs]$ \citep{feller2017principal}.
That is, the principal stratum is unrelated to control potential outcomes conditional on covariates.
Principal ignorability \eqref{eq:pi} is reminiscent of ignorability assumptions typical in observational studies. % \citep[e.g.][]{rosenbaum2010design}.
% Both require a rich set of covariates and background knowledge of the determinants of $S$; 
In particular, an unobserved covariate $u$ that is correlated with both $\st$ and $Y_C$ would invalidate \eqref{eq:pi}.

@@

Principal score weighting does not require any distributional assumptions and (after estimating principal scores) is very easy to implement.
On the other hand, the principal ignorability assumption is strong and restrictive.

\citet{feller2017principal} recommends the case-resampling bootstrap to estimate the sampling variances of $\hat{\mu}_{Cw}^0$, $\hat{\mu}_{Cw}^1$, and their associated principal effect estimators.

@@

We conducted a simulation study to compare the performance of GEEPERs compared to a Bayesian mixture model and to a principal score weighting estimator, in both favorable and unfavorable circumstances.
The study was designed to answer three sets of overarching questions.
First, how does the performance of the GEEPERs estimator vary under different conditions, including sample size, the extent to which $\bx$ predicts $\st$, whether and how the principal effects themselves vary, and when important interactions are omitted from the outcome model.
Second, how does GEEPERs compare with mixture modeling? Is it competitive in circumstances favorable to both techniques? Does it avoid the pitfalls of mixture modeling when the parametric assumptions of the normal mixture model fail?
Third, are there ways or circumstances when the principal score weighting estimator outperforms GEEPERs even though principal ignorability does not hold?

@@

In each run of the simulation, we simulated three independent standard normal covariates, $x_k$ $k=1,\dots,3$. However, only the first two covariates were ``observed,'' i.e. were included in the analysis model.
Given the covariates, the (true) principal scores were defined as
\begin{equation*}
  \ppi=logit^{-1}\left[\alpha(x_{1i}-x_{2i}+x_{3i})\right]
\end{equation*}
where $\alpha$ is a manipulated factor.
When $\alpha$ was higher, $\st$ was more easily predicted by covariates; when $\alpha=0$, Assumption \ref{ass:vps} was violated, i.e. $Var(\pp)=0$.
Principal stratum $\sti\in \{0,1\}$ was simulated as $S_i\sim Bern(\ppi)$

Potential and observed outcomes were generated as 
\begin{equation}\label{eq:y-sim}
Y_i=\beta_1\sti+\beta_3Z_i\sti+(\gamma_1+\gamma_2\sti+\gamma_3Z_i)(x_{1i}+x_{2i})+\frac{1}{\sqrt{6}}x_{3i}+\epsilon_i%\\
\end{equation}
%   \begin{split}
%   \yci&=\beta_1\sti+(\gamma_1+\gamma_2\sti)(x_{1i}+x_{2i})+\frac{1}{\sqrt{6}}x_{3i}+\epsilon_i\\
%   \yti&=0.3\sti+(\gamma_1+\gamma_2\sti+\gamma_3Z_i)(x_{1i}+x_{2i})+\frac{1}{\sqrt{6}}x_{3i}+\epsilon_i\\
%   Y_i&=Z_i\yti+(1-Z_i)\yci
%   \end{split}
% \end{equation}
with $\beta_3=0.3-\beta_1$, $\EE[\epsilon_i]=0$, and $Var(\epsilon_i)=1/2$. When $\gamma_3=0$ (i.e. no interaction between $Z$ and covariates), there is no average treatment effect in the $\st=0$ stratum.

@@

\textbf{\emph{Sample Size per condition $\bm{n}$:}} The guarantees of M-estimation are all asymptotic, so we examined the behavior of our GEEPERs at a range of sample sizes to determine its finite-sample properties.

@@

Following \citetsupp[][p. 373]{carroll2006measurement}, we %separate the estimating equations $\Psi$ into $\phi$ and $\psi$, estimating equations for models \eqref{
can decompose the matrices into diagonal elements 
\begin{equation*}
    \begin{split}
        A_{1,1}&=\sum_i \partial \Omega_i/\partial \bm{\alpha}|_{\bm{\alpha}=\bm{\hat{\alpha}}}\\
        A_{2,2}&=\sum_i\partial\Psi_i/\partial \bm{\beta}|_{\bm{\beta}=\bm{\hat{\beta}}}\\
        B_{1,1}&=\sum_i\Omega_i(\bm{\hat\alpha})\Omega_i(\bm{\hat\alpha})'\\
        B_{2,2}&=\sum_i \Psi_i(\bm{\hat\alpha},\bm{\hat\beta})\Psi_i(\bm{\hat\alpha},\bm{\hat\beta})'
    \end{split}
\end{equation*}
 that pertain to the parameter sets $\bm{\alpha}$ and $\bm{\beta}$ and the estimating equations for models \eqref{eq:psMod} and  \eqref{eq:regression}, respectively, and 
 \begin{equation*}
     \begin{split}
         A_{21}&=\sum_i\partial\Psi_i/\partial \bm{\alpha}|_{\bm{\alpha}=\bm{\hat{\alpha}}}\\
         B_{12}=B_{21}'&=\sum_i \Omega_i(\bm{\hat{\alpha}})\Psi_i(\bm{\hat{\alpha}},\bm{\hat{\beta}})'
     \end{split}
 \end{equation*}
 which capture to the dependence of model \eqref{eq:regression} on the parameters $\bm{\alpha}$ from \eqref{eq:psMod} and the covariance between the estimating equations of the two models. 
The sub-matrix $A_{12}=\sum_i \partial \Omega_i/\partial \bm{\beta}=0$, since \eqref{eq:psMod} does not depend on $\bm{\beta}$.

@@

\bibliographystylesupp{plainnat}
\bibliographysupp{MOM}
